GEE models

### 2016-05-17 doc/manuscript.Rmd
Updated manuscript as per Tony's edits

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output:
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 7
        fig_caption: true
        keep_md: true
---

<!--

TODO:
- Use pct for PLSDA

@@

Word count: 3569 / 4000

@@

<!-- Reference limit / 40 -->

# Abstract
249 / 250 words

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acids (NEFA) on changes in insulin sensitivity and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and who had completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values.
Insulin sensitivity was assessed using 1/HOMA-IR and the Matsuda index (ISI),
while beta-cell function was assessed using the Insulinogenic index over HOMA-IR
(IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized
estimating equations was used, adjusting for time, waist, sex, ethnicity, age,
ALT, alcohol intake, physical activity, and family history of diabetes. NEFA
were analyzed as both a concentration and as a proportion of the total fraction.

Results: Insulin sensitivity and beta-cell function declined by 14-27% while
BMI did not change over the 6-years. Total NEFA, 16:0, 18:1n-9, and 18:2n-6 as
concentrations had negative associations with IGI/IR and ISSI-2. These four NEFA
variables had similar magnitudes of association, predicting between 4-8% lower
IGI/IR and ISSI-2 over the 6 years. No associations were seen for 1/HOMA-IR
or ISI, nor were any associations seen for the mol% data.

Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
6 years. Our results suggest that association with beta-cell function is due to
the absolute size of the serum NEFA fraction, rather than the specific fatty acid
composition.

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TAG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
- [X] Replace TAG with BaseTAG in the QIC (or use both)
- [X] Fix time axis labels on LCMM plot
- Describe LCMM and PLS-DA better in plots

@@

```{r knit_setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'))
```

@@

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have compared the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in the research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our research objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models (M5-M8) were comparable in performance, so we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we a priori
decided to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TAG is a risk factor
for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
between NEFA and the outcomes. To determine the role of TAG in the
association between NEFA and the outcomes, TAG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. As such, after running the GEE models for the
longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
used to identify the patterns of NEFA composition against beta-cell function
(using ISSI-2, which had most of the associations in the GEE models) and for
those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
mixed models (LCMM) were used to extract latent trajectories of beta-cell
function (ISSI-2) over the 6 years.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].

@@

For this study, there were 
`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables had a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this analysis of the PROMISE cohort (n=`r percent.change.outcomes$n`).
Consistent with this decline, there were `r dm.convert`
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.

`r cite_f('nefaDist')` shows the compositional distribution of NEFA in the study
participants. Four individual NEFA made up the vast majority 
(`r sum(percent.nefa$pct)`%) of the total NEFA fraction. The largest contributors
were `r percent.nefa$c[1]`, `r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and 
`r percent.nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefaDist')`. 
All individual NEFA as well as the total fraction had
correlations that ranged from weak to null (r<0.3) with the basic
characteristics (see `r cite_sf('corr')`).

@@

Several associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for the concentration compared to the
mol% data. Full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
had negative associations between
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD decrease
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
(data not shown).
There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (all p>0.15). Raw values from the GEE models are shown in 
`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function, however, those individuals 
classified into the low beta-cell function group tended to have patterns of NEFA 
composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status over the 6 years as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants
were classified correctly based on dysglycemia status, examining the clustering
of the NEFA composition showed poor discrimination between groups (see 
`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA concentrations independently predicted
lower beta-cell function after 6 years. While we found negative associations with
palmitic acid (16:0), oleic acid (18:1n-9), and
linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
associations were seen
for the proportion of these fatty acids and further clustering analysis found no
predictive ability of these NEFA on beta-cell function or dysglycemia. These
observations suggest that the absolute size of the total NEFA fraction, rather than
the specific composition of any individual NEFA, likely influences the
pathogenesis of diabetes.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
total NEFA had a negative association with insulin sensitivity and a
particularly strong negative association with beta-cell function [@Johns2014a].
Experimentally, several potential mechanisms have been elucidated for the role
of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
through endoplasmic reticulum stress, formation of ceramides, and generation of
nitric oxide, as well as impairment of proinsulin production and mitochrondrial
function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
The present analysis is the first study to our knowledge to examine the 
longitudinal association of a broad spectrum of individual NEFA on
beta-cell function in a large cohort. We found that there was a strong signal of
higher total NEFA, palmitic acid, oleic acid, and linoleic acid as
concentrations with lower beta-cell function. However, in modeling these fatty
acids as a mol% and using novel clustering analysis approaches, no individual or
specific fatty acids in the NEFA fraction strongly predicted lower beta-cell
function, insulin sensitivity, or dysglycemia status. Taken together, these
results suggest that it is the absolute size of the circulating NEFA fraction,
irrespective of any specific composition of fatty acids, that is
responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.

Biologically, within a free-living population chronic elevation of NEFA may be 
mediating it's association through TAG. In normal metabolism, NEFA enters the 
liver and assists in the production of TAG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TAG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating it's association
through TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are some possible explanations for these findings. Previous studies done *in
vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.

There are two other possible explanations for our null findings for the insulin
sensitivity measures. First, there may be differences in physiology between
fasting and postprandial NEFA kinetics. For instance, some experimental studies
using clamp protocols found that fasting NEFA was a weak predictor of insulin
sensitivity compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
NEFA spillover into the blood and a subsequent increase in circulating NEFA
[@Almandoz2013a], which may be more metabolically active given postprandial 
activity. Second, the null findings may be due to the high risk population
examined in PROMISE. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat mode advanced
stage in the pathogenesis of diabetes.

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output: fost::diabetes_care_manuscript
---

@@

Word count: 3577 / 4000

@@

<!-- Reference limit 41 / 40 -->

# Abstract
248 / 250 words

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and who had completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values.
IS was assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function
was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, age, ALT, alcohol intake,
physical activity, and family history of diabetes. NEFA were analyzed as both
concentrations and proportions (mol%) of the total fraction.

Results: IS and beta-cell function declined by 14-27% while BMI
did not significantly change over 6 years of follow-up. Total NEFA, 16:0,
18:1n-9, and 18:2n-6 as concentrations had significant negative associations
with IGI/IR and ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the
6 years. However, individual NEFA modeled as mol% showed no associations with
these outcomes. Total and individual NEFA were not associated with IS measures.

Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell
function over 6 years. Our results suggest that the association with beta-cell
function is due to the absolute size of the serum NEFA fraction, rather than the
specific fatty acid composition.

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TAG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
- [X] Replace TAG with BaseTAG in the QIC (or use both)
- [X] Fix time axis labels on LCMM plot
- Describe LCMM and PLS-DA better in plots
- [X] Use pct for PLSDA
- [X] Explained variation in X and in Y for PLS-DA
- Supplemental only file (separate from manuscript).. maybe?

@@

```{r knit_setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'), dpi = 180)
```

@@

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see [@Johnston2016a] for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last four models (M5-M8) were comparable in performance, therefore we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we decided *a priori*
to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TAG is a risk factor
for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
between NEFA and the outcomes. To determine the role of TAG in the
association between NEFA and the outcomes, TAG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. As such, after running the GEE models for the
longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
used to identify the patterns of NEFA composition against beta-cell function
(using ISSI-2, which had the majority of the significant associations in the GEE models) and for
those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
mixed models (LCMM) were used to extract latent trajectories of beta-cell
function (ISSI-2) over the 6 years.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{add link before
submission}}. Results were considered statistically significance at p<0.05,
after adjusting for multiple testing using the Benjamini-Hochberg False
Discovery Rate [@Benjamini1995a].

@@

For this study, there were 
`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables showed a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this analysis of the PROMISE cohort.
Consistent with this decline, there were `r dm.convert`
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.

`r cite_f('nefaDist')` shows the compositional distribution of NEFA in the study
participants. Four individual NEFA made up the vast majority 
(`r sum(percent.nefa$pct)`%) of the total NEFA fraction. The largest contributors
were `r percent.nefa$c[1]`, `r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and 
`r percent.nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefaDist')`. 
All individual NEFA as well as the total fraction had
correlations that ranged from weak to null (r<0.3) with participant
characteristics (see `r cite_sf('corr')`).

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
Full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
had negative associations with
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as concentrations, none of the NEFA variables modeled as mol% were associated with the
outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
(data not shown).
There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (all p>0.15). Raw values from the GEE models are shown in 
`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function, however, those individuals 
classified into the low beta-cell function group tended to have patterns of NEFA 
composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status over the 6 years as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants 
were classified correctly based on dysglycemia status, this was almost entirely
due to classifying correctly into the non-dysglycemia status. Examining the
clustering of the NEFA composition showed poor discrimination between groups
(see `r cite_sf('plsda_groups_dysgly')` and 
`r cite_sf('plsda_loadings_dysgly')`).

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA concentrations independently predicted
lower beta-cell function after 6 years. While we found negative associations with
palmitic acid (16:0), oleic acid (18:1n-9), and
linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
associations were seen for the proportion of these fatty acids and further
clustering analysis found no predictive ability of these individual NEFA on
beta-cell function or dysglycemia. These observations suggest that the absolute
size of the total NEFA fraction, rather than its specific composition, likely
influences the pathogenesis of diabetes.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
total NEFA had a negative association with insulin sensitivity and a
particularly strong negative association with beta-cell function [@Johns2014a].
Experimentally, several potential mechanisms have been elucidated for the role
of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
through endoplasmic reticulum stress, formation of ceramides, and generation of
nitric oxide, as well as impairment of proinsulin production and mitochondrial
function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
The present analysis is the first study to our knowledge to examine the 
longitudinal association of a broad spectrum of individual NEFA on
beta-cell function in a large cohort. We found that there was a strong signal of
higher total NEFA, palmitic acid, oleic acid, and linoleic acid modeled as
concentrations with lower beta-cell function. However, in modeling these fatty
acids as a mol% and using novel clustering analysis approaches, no specific
fatty acids in the NEFA fraction strongly predicted lower beta-cell
function, insulin sensitivity, or dysglycemia status. Taken together, these
results suggest that it is the absolute size of the circulating NEFA fraction,
irrespective of any specific composition of fatty acids, that is
responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating it's association with metabolic outcomes through TAG. In normal
metabolism, NEFA enters the
liver and assists in the production of TAG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such, higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TAG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating it's association
with beta-cell dysfunction through higher TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are several possible explanations for these findings. Previous studies
done *in vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and previous observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in free-living environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA may be a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.

There are two other possible explanations for our null findings for the insulin
sensitivity measures. First, there may be differences in physiology between
fasting and postprandial NEFA kinetics. For instance, some experimental studies
using clamp protocols found that fasting NEFA was a weak predictor of insulin
sensitivity compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
NEFA spillover into the blood and a subsequent increase in circulating NEFA
[@Almandoz2013a], which may be more metabolically active given postprandial 
activity. Second, the null findings may be due to the high risk population
examined in PROMISE. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.
